A Simplified Spectral Deferred Correction Method for Coupling Hydrodynamics with Reaction Networks and Nuclear Statistical Equilibrium

### 2020-10-22 simplified_sdc_intro/paper.tex
add some tables
fixed typos

\newcommand{\nablab}{{\mathbf{\nabla}}}
\newcommand{\Ub}{\mathbf{U}}
\newcommand{\gb}{\mathbf{g}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\Sdot}{\dot{S}}
\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

@@

\newcommand{\nse}[1]{{\mathrm{NSE}( #1 )}}

\newcommand{\rhonse}{{\rho_\mathrm{nse}}}
\newcommand{\tnse}{{T_\mathrm{nse}}}
\newcommand{\Anse}{{A_\mathrm{nse}}}
\newcommand{\Bnse}{{B_\mathrm{nse}}}

\newcommand{\out}{{\rm out}}
\newcommand{\inp}{{\rm in}}

@@

A particular difficult phase of evolution to model is the nuclear
statistic equilibrium that sets in with helium burning for
temperatures in excess of $4\times 10^9~\mathrm{K}$.  Physically, the
forward and reverse rates should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  This large change in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep, but this makes simulations prohibitively expensive.
Sometimes the burning is simply halted on a zone-by-zone basis when
NSE conditions are reached (e.g., as in \citealt{hedet}).
Alternately, at high temperatures, a reaction network can be replaced
with a table of NSE abundances and the zone's composition set through
table look-ups (e.g.\ \citealt{ma:2013}) \MarginPar{need more here}

In \cite{castro_sdc}, we developed second- and fourth-order accurate
methods in space and time for coupling hydrodynamics and nuclear
reaction networks based on spectral deferred corrections (SDC)
methods, and demonstrated the method of a variety of test problems.
The approach presented here is considerably simpler, reusing the main
CTU hydrodynamics construction and a largely similar ODE integration
scheme, making this method easier to add to existing simulation codes.
However, it is restricted to second-order accuracy overall.  We term
this algorithm the ``simplified SDC method''.  We use the
\castro\ hydrodynamics code \cite{castro} for all of our numerical
experiments, and all of the code to reproduce the results in this
paper are in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  We'll
focus on one-dimensional flow for this description, and the system
appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho \alpha_l \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity,
$u$, specific total energy, $E$, and nuclear species mass fractions,
$X_k$, and auxiliary composition variables, $a_l$.  The mass fractions
are constrained to sum to 1, $\sum_k X_k = 1$, but no such constraint
exists on the auxiliary variables.  The specific total energy relates
to the specific internal energy as $E = e + u^2/2$.  The corresponding
fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\
   \rho X_k u \\
   \rho \alpha_l u \\
   \rho u u + p \\
   \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state.  For a system
where the composition is completely specified by the mass fractions, $X_k$, the equation
of state would take the form:
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}
However, as we'll see shortly, when we use the NSE table, we are using
the results of a much larger network then can be represented by the
mass fractions, and in this case, we rely on the auxiliary composition
variables, $\alpha_l$, to describe the state of the composition, and
our EOS has the form:
\begin{equation}
p = p(\rho, \alpha_l, e)
\end{equation}

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc)
\end{equation}
 The reactive
sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\
     \rho \omegadot(X_k) \\
     \rho \omegadot(\alpha_l) \\
     0 \\
     \rho \dot{S}
  \end{array} \right )
\end{equation}
where $\omegadot(X_k)$ is the creation rate for species $k$, $\omegadot(\alpha_l)$ is the creation rate for
auxiliary composition variable $l$, and $\dot{S}$
is the energy generation rate per unit mass.

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  \alpha_l \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sq
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{cccccc}
    u & 0 & 0 & \rho & 0 & 0 \\
    0 & u & 0 & 0    & 0 & 0 \\
    0 & 0 & u & 0    & 0 & 0 \\
    0 & 0 & 0 & u    & 1/\rho & 0 \\
    0 & 0 & 0 & \Gamma_1 p & u & 0 \\
    0 & 0 & 0 & \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $\Gamma_1$ is an adiabatic index, $\Gamma_1 = d\log p/d\log\rho
|_s$.  Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sq$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sq = \Sqhydro + \Rb(q)
\end{equation}
with
\begin{equation}
\label{eq:prim_sources}
\Rb(q) = \left ( \begin{array}{c}
     0 \\
     \omegadot(X_k) \\
     \omegadot(\alpha_l) \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.

We use the \castro\ hydrodynamics code~\cite{castro}, together with
the corner transport upwind piecewise parabolic method
hydrodynamics~\citep{millercolella:2002}.  This is a finite-volume
method that uses characteristic tracing to predict a time-centered
flux through the interfaces of the grid zones.  For pure
hydrodynamics, these time-centered fluxes result in second-order
accurate time integration.  With reactions, we want to couple
hydrodynamics and the reaction sources to second order.  Nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting is traditionally employed
here, and we compare to it in this paper.

@@

In the absence of advective terms, our reaction system appears as just
$d\Uc/dt = \Rb(\Uc)$, or:
\begin{align}
\odt{\rho} & = 0 \\
\odt{(\rho X_k)} &= \rho \omegadot(X_k) \\
\odt{(\rho \alpha_l)} &= \rho \omegadot(\alpha_l) \\
\odt{(\rho u)} &= 0 \\
\odt{(\rho E)} &= \rho \Sdot
\end{align}
We can write the energy equation as:
\begin{equation}
\odt{(\rho E)} = \odt{(\rho e)} + \odt{K} = \rho \Sdot
\end{equation}
where $K$ is the kinetic energy, $K = |u|^2/2$.  Since the velocity is
unchanged by reactions, this means that
\begin{equation}
\odt{(\rho e)} = \rho \odt{e} = \rho \Sdot
\end{equation}
The reaction rates are typically expressed as $\omegadot_k(\rho, T,
X_k)$, when we evolve this system.  This requires an EOS inversion to get $T$ from $e$ each
time we need to evaluate the reactive terms.

Normally we want to integrate temperature temperature.  We can write the
energy equation as:
\begin{equation}
\rho \odt{e} = \rho \left [ \underbrace{\left . \frac{\partial e}{\partial T} \right |_\rho}_{c_v}
                            \odt{T} +
                            \left . \frac{\partial e}{\partial \rho} \right |_T
                            \cancelto{0}{\odt{\rho}} \;\; \right ] = \rho \Sdot
\end{equation}
reducing to
\begin{equation}
\label{eq:strang:T}
\odt{T} = \frac{1}{c_v} \Sdot
\end{equation}
We can alternately derive this from an enthalpy equation, which is the
correct formulation when we are at constant pressure, and we get the
same temperature equation but with $c_v$ replaced by $c_p$.
Technically this requires an EOS call to get $c_x$, but if the
specific heat varies slowly, we can freeze it during the reaction
solve or use a Taylor expansion in $T$ to capture slow evolution.

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling) as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep:
\begin{align}
\label{eq:dt:nuce}
\Delta t &\le f_e\, \min_{i} \left\{\frac{e_{i}}{\dot{S}_{i}}\right\} \\
%
\label{eq:dt:nucX}
\Delta t &\le f_X\, \min_{i} \left\{\min_{k,X_k > \epsilon_X}\frac{{X_k}_{i}}{{\omegadot(X_k)_i}}\right\}
\end{align}
where $i$ is the zone index and $f_e$ and $f_X$ are runtime parameters
used to control the allowed change, and only species for which $X_k > \epsilon_X$ are considered.

\castro\ has the ability to reject a timestep if it detects a failure
and retry with smaller timesteps (subcycling to make up the original
required timestep).  Among the conditions that can trigger this are
density failing below zero during advection, the ODE integration
failing to converge in the implicit solve, or violation of one of the
timestep limiters during the step.  This means that equations
\ref{eq:dt:nuce} and \ref{eq:dt:nucX} are not reactive, but instead
guaranteed to be met throughout the simulation, because the step is
rejected if they are violated.  This contrasts to similar approaches
used in other codes where conditions \ref{eq:dt:nuce} and
\ref{eq:dt:nucX} are used to restrict the next timestep size, but the
update over the current step is kept, which already violated the
constraints. \MarginPar{would be nice to have a reference here} The retry mechanism in \castro\ works with both the
Strang and simplified-SDC integration scheme.

@@

We start by writing our update as:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt
\end{equation}
For the simplified SDC method we explore here, we will make the
advective term piecewise constant in time, using the value at the
midpoint in time, to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
This is the approach first shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  A compressible version, for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below.

@@

\label{sec:initialization}

  \begin{itemize}
  \item We need an approximation to how much the reactions alone
    changed the primitive variable state over the timestep, which we
    will denote $\Ic_q$.  Since we do not have any information about
    the current timestep in the first iteration,
    we use the value from the last iteration of the previous timestep:
    \begin{equation}
      \Ic^{n+1/2,(-1)}_{\qb} = \Ic^{n-1/2,(\smax-1)}_{\qb}
    \end{equation}
  %% \item Solve the Poisson problem for the initial gravitational potential:
  %%   \begin{equation}
  %%     \nabla^2 \Phi^n = 4\pi G \rho^n
  %%   \end{equation}

@@

    We now seek the $\Ic$'s that capture the effect of just the
    reaction sources on the state variables for the next iteration.
    For the conserved quantities, these would simply be: \MarginPar{fix signs on A}
    \begin{equation}
      \Ic^{(k)}_{\Uc} = \frac{\Uc^{n+1,(k)} - \Uc^n}{\Delta t} - \Adv{\Uc}^{n+1/2,(k)}
    \end{equation}
    However, for our primitive variables, which are used in the
    prediction, we need to construct the required source terms more
    carefully.  We want:
    \begin{equation}
      \label{eq:Iq}
      \Ic^{(k)}_{\qb} = \frac{\qb^{n+1,(k)} - \qb^n}{\Delta t} - \Adv{\qb}^{n+1/2,(k)}
    \end{equation}
    but we need the advective update for $\qb$, which we have not
    constructed.  We note that $\Ic^{(k)}_\qb$ is an approximation to the integral of 
    Eq.~\ref{eq:prim_sources} over the timestep.  Additionally, we cannot simply use the equation of
    state on $\Ic^{(k)}_{\Uc}$ since this is a time-derivative and
    does not represent a well-defined state in itself.  Instead, we
    derive $\Ic^{(k)}_{\qb}$ via a multi-step process.  We first find
    the conservative state as if it were updated only with advection:
    \begin{equation}
      \Uc^\star = \Uc^n + \Delta t \Adv{\Uc}^{n+1/2,(k)}
    \end{equation}
    and then construct the corresponding primitive variable state via an algebraic transform,
    $\Uc^\star \rightarrow \qb^\star$.
    This allows us to define the advective update for $\qb$ as:
    \begin{equation}
      \Adv{\qb}^{n+1/2,(k)} = \frac{\qb^\star - \qb^n}{\Delta t}
    \end{equation}
    Defining the primitive state corresponding to the fully-updated
    conserved state via an algebraic transform, $\Uc^{n+1,(k)}
    \rightarrow \qb^{n+1,(k)}$, we can construct $\Ic^{(k)}_{\qb}$ as given
    in Eq.~\ref{eq:Iq}.
    Putting all of this together, we see:
    \begin{equation}
      \Ic^{(k)}_{\qb} = \frac{\qb^{n+1,(k)} - \qb^\star}{\Delta t}
    \end{equation}

@@

For this study we will use the 19 nuceli network containing
\isot{H}{1}, \isot{He}{3}, \isot{He}{4}, \isot{C}{12}, \isot{N}{14},
\isot{O}{16}, \isot{Ne}{20}, \isot{Mg}{24}, \isot{Si}{28},
\isot{S}{32}, \isot{Ar}{36}, \isot{Ca}{40}, \isot{Ti}{44},
\isot{Cr}{48}, \isot{Fe}{52}, \isot{Fe}{54}, \isot{Ni}{56}, protons
(from photodisintegration), and neutrons.  This is based on the
``aprox19'' network from \cite{aprox19} and originally described in
\cite{Kepler}.  We combine this with a table that gives the nuclear
statistical equilibrium (NSE) abundances in regions where the system
is in NSE.  The NSE table was generated using a 127 nuclei reaction
network and is the same as described in \cite{ma:2013}.  In our
simulations, we carry all 19 isotopes in the main network in each zone
and advect them in the hydrodynamics portion of the algorithm.  The
composition of the larger 127 nuclei network is mapped into the 19
isotopes we carry according to table~\ref{table:nuclei}.

The NSE table requires:
\begin{equation}
  Y_e = \sum_k \frac{Z_k X_k}{A_k}
\end{equation}
(where $A_k$ and $Z_k$ are the atomic weight and atomic number of nucleus $k$) and provides
\begin{align}
\bar{A} &= \left [ \sum_k \frac{X_k}{A_k} \right ]^{-1} \\
\left (\frac{B}{A} \right ) &= \sum_k \frac{B_k X_k}{A_k}
\end{align}
where $B_k$ is the binding energy of nucleus $k$.  In our simulations,
we store these 3 quantities as auxiliary data that is carried along
with the rest of the fluid state in each zone.  The table also returns
values of the mass fractions mapped onto the 19-isotopes we carry, $X_k$, and the time-derivative of $Y_e$, $dY_e/dt$.
We'll use the notation:
\begin{equation}
\nse{\rho,T,Y_e} \rightarrow \bar{A}, X_k, (B/A), dY_e/dt
\end{equation}
to represent the NSE table call and its inputs and outputs.

>>>>

\newcommand{\nablab}{{\mathbf{\nabla}}}
\newcommand{\Ub}{\mathbf{U}}
\newcommand{\gb}{\mathbf{g}}
\newcommand{\omegadot}{\dot{\omega}}
\newcommand{\Sdot}{\dot{S}}
\newcommand{\ddx}[1]{{\frac{{\partial#1}}{\partial x}}}
\newcommand{\ddt}[1]{{\frac{{\partial#1}}{\partial t}}}
\newcommand{\odt}[1]{{\frac{{d#1}}{dt}}}
\newcommand{\divg}[1]{{\nablab \cdot \left (#1\right)}}
\newcommand{\inp}{{\mathrm{input}}}
\newcommand{\dedr}{\left . {\partial{}e}/{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedrd}{\left . \frac{\partial{}e}{\partial\rho}\right |_{T, X_k}}
\newcommand{\dedX}{\left . {\partial{}e}/{\partial{}X_k} \right |_{\rho, T}}
\newcommand{\dedXd}{\left . \frac{\partial{}e}{\partial{}X_k} \right |_{\rho, T, X_{j,j\ne k}}}
\newcommand{\dedT}{\left . {\partial{}e}/{\partial{}T} \right |_{\rho,X_k}}
\newcommand{\dedTd}{\left . \frac{\partial{}e}{\partial{}T} \right |_{\rho,X_k}}

@@

@@

A particularly difficult phase of evolution to model is the nuclear
statistical equilibrium that sets in with helium burning for
temperatures in excess of $4\times 10^9~\mathrm{K}$.  Physically, the
forward and reverse rates of reaction should balance leading to an equilibrium.
With operator splitting, an NSE region will have a large positive flow
through the network in a zone in one step followed by a large negative
flow over the next timestep, as the code struggles to produce an
equilibrium.  This large change in abundances (and large alternately
positive and negative energy generation rates) can be a challenge for
a code.  The easiest way to improve the coupling is to cut the
timestep, but this makes simulations prohibitively expensive.
Sometimes the burning is simply halted on a zone-by-zone basis when
NSE conditions are reached (e.g., as in \citealt{hedet}).
Alternately, at high temperatures, a reaction network can be replaced
with a table of NSE abundances and the zone's composition set through
table look-ups (e.g.\ \citealt{ma:2013}) \MarginPar{need more here}

In \cite{castro_sdc}, we developed second- and fourth-order accurate
methods in space and time for coupling hydrodynamics and nuclear
reaction networks based on spectral deferred corrections (SDC)
methods, and demonstrated the method in a variety of test problems.
The approach presented here is considerably simpler, reusing the main
CTU hydrodynamics construction and a largely similar ODE integration
scheme, making this method easier to add to existing simulation codes.
However, it is restricted to second-order accuracy overall.  We term
this algorithm the ``simplified SDC method''.  We use the
\castro\ hydrodynamics code \cite{castro} for all of our numerical
experiments, and all of the code to reproduce the results in this
paper is in the \castro\ github
repository\footnote{\url{https://github.com/amrex-astro/Castro/}}.

@@

We solve the Euler equations for compressible, reacting flow.  We'll
focus on one-dimensional flow for this description, and the system
appears as:
\begin{equation}
\frac{\partial \Uc_t}{\partial t}  + \frac{\partial \Fb^\xv (\Uc)}{\partial x} = \Sc(\Uc)
\end{equation}
where
\begin{equation}
\Uc = \left ( \begin{array}{c}
           \rho \\
           \rho X_k \\
           \rho u \\
           \rho E \end{array}\right )
\end{equation}
are the conserved fluid quantities: mass density, $\rho$, velocity, $u$,
specific total energy, $E$, and nuclear
species mass fractions, $X_k$.  The specific total energy relates to
the specific internal energy as $E = e + u^2/2$.  The
corresponding fluxes are
\begin{equation}
\Fb^\xv (\Uc) = \left ( \begin{array}{c}
   \rho u \\ \rho X_k u \\ \rho u u + p \\  \rho u E + u p
   \end{array} \right )
\end{equation}
Here the pressure, $p$, enters, and is found via the equation of state,
\begin{equation}
p = p(\rho, X_k, e)
\end{equation}

We decompose the source terms, $\Sc(\Uc)$, into
hydrodynamical sources, $\Hb$ (like gravity), and reactive sources,
$\Rb$,
\begin{equation}
  \Sc(\Uc) = \Hb(\Uc) + \Rb(\Uc)
\end{equation}
 The reactive
sources take the form:
\begin{equation}
  \Rb(\Uc) = \left ( \begin{array}{c}
     0 \\ \rho \omegadot_k \\ 0 \\ \rho \dot{S}
  \end{array} \right )
\end{equation}
where $\omegadot_k$ is the creation rate for species $k$ and $\dot{S}$
is the energy generation rate per unit mass.

Sometimes it is preferable to work with the primitive variables,
\begin{equation}
\qb = \left ( \begin{array}{c}
  \rho \\
  X_k \\
  u \\
  p \\
  (\rho e) \\
\end{array} \right )
\end{equation}
Here, the system appears
as:
\begin{equation}
\qb_t + \Ab^\xv(\qb) \qb_x  = \Sq
\end{equation}
with the matrix $\Ab^\xv$ giving the coefficients of the spatial derivatives
of the primitive variables:
\begin{equation}
\Ab^\xv(\qb) = \left ( \begin{array}{ccccc}
    u & 0 & \rho & 0 & 0 \\
    0 & u & 0    & 0 & 0 \\
    0 & 0 & u    & 1/\rho & 0 \\
    0 & 0 & \Gamma_1 p & u & 0 \\
    0 & 0 & \rho h & 0 & u
  \end{array} \right )
\end{equation}
where $\Gamma_1$ is an adiabatic index, $\Gamma_1 = d\log p/d\log\rho
|_s$.  Note, the primitive state has two thermodynamic quantities, $p$
and $(\rho e)$, to more efficiently handle the general equation of
state in the Riemann solver, as described in \citet{castro}, but
alternate formulations are possible \citep{colellaglaz:1985}.
The source term vector, $\Sq$, can again be decomposed into hydrodynamic
sources (now in terms of the primitive variables) and reaction terms,
\begin{equation}
  \Sq = \Sqhydro + \Rb(q)
\end{equation}
with
\begin{equation}
\Rb(q) = \left ( \begin{array}{c}
     0 \\
     \omegadot_k \\
     0 \\
     \Gamma_1 p \sigma \Sdot \\
     \rho \Sdot
   \end{array} \right )
\end{equation}
where
\begin{equation}
\sigma \equiv \frac{\partial p/\partial T |_\rho}{\rho c_p \partial p/\partial \rho |_T}
\end{equation}
and $c_p$ is the specific heat at constant pressure, $c_p = \partial
h/\partial T |_p$.  A derivation of this source for the pressure
equation can be found in \cite{ABNZ:III}.  We note that this source is
algebraically identical to that shown in Eq.~25 of \cite{castro}.

We use the \castro\ hydrodynamics code~\cite{castro}, together with
corner transport upwind piecewise parabolic method
hydrodynamics~\citep{millercolella:2002}.  This is a finite-volume
method that uses characteristic tracing to predict a time-centered
flux through the interfaces of the grid zones.  For pure
hydrodynamics, these time-centered fluxes result in second-order
accurate time integration.  With reactions, we want to couple
hydrodynamics and the reaction sources to second order.  Nuclear
reaction sources are stiff, and need to be integrated using implicit
methods for stabilty.  Operator splitting is traditionally employed
here, and we compare to it in this paper.

@@

In the absence of advective terms, our reaction system appears as just
$d\Uc/dt = \Rb(\Uc)$, or:
\begin{align}
\odt{\rho} & = 0 \\
\odt{(\rho X_k)} &= \rho \omegadot_k \\
\odt{(\rho u)} &= 0 \\
\odt{(\rho E)} &= \rho \Sdot
\end{align}
We can write the energy equation as:
\begin{equation}
\odt{(\rho E)} = \odt{(\rho e)} + \odt{K} = \rho \Sdot
\end{equation}
where $K$ is the kinetic energy, $K = |u|^2/2$.  Since the velocity is
unchanged by reactions, this means that
\begin{equation}
\odt{(\rho e)} = \rho \odt{e} = \rho \Sdot
\end{equation}
The reaction rates are typically expressed as $\omegadot_k(\rho, T,
X_k)$, when we evolve this system.  This requires an EOS inversion to get $T$ from $e$ each
time we need to evaluate the reactive terms.

Normally we want to integrate the temperature.  We can write the
energy equation as:
\begin{equation}
\rho \odt{e} = \rho \left [ \underbrace{\left . \frac{\partial e}{\partial T} \right |_\rho}_{c_v}
                            \odt{T} +
                            \left . \frac{\partial e}{\partial \rho} \right |_T
                            \cancelto{0}{\odt{\rho}} \;\; \right ] = \rho \Sdot
\end{equation}
reducing to
\begin{equation}
\label{eq:strang:T}
\odt{T} = \frac{1}{c_v} \Sdot
\end{equation}
We can alternately derive this from an enthalpy equation, which is the
correct formulation when we are at constant pressure, and we get the
same temperature equation but with $c_v$ replaced by $c_p$.
Technically this requires an EOS call to get $c_x$, but if the
specific heat varies slowly, we can freeze it during the reaction
solve or use a Taylor expansion in $T$ to capture slow evolution.

@@

Since this method is based off of the CTU hydrodynamics scheme, it
benefits from the larger timestep that method can take (when done with
full corner coupling) as compared to a method-of-lines approach (see
\citealt{ppmunsplit}).  In addition to the standard CFL timestep
limiter for explicit hydrodynamics, \castro\ can also enforce
timestep limiters based on the energy generation or abundance changes
over a timestep:
\begin{align}
\label{eq:dt:nuce}
\Delta t &\le f_e\, \min_{i} \left\{\frac{e_{i}}{\dot{e}_{i}}\right\} \\
%
\label{eq:dt:nucX}
\Delta t &\le f_X\, \min_{i} \left\{\min_{k,X_k > \epsilon_X}\frac{{X_k}_{i}}{{\omegadot_{k,i}}}\right\}
\end{align}
where $i$ is the zone index and $f_e$ and $f_X$ are runtime parameters
used to control the allowed change, and only species for which $X_k > \epsilon_X$ are considered.

\castro\ has the ability to reject a timestep if it detects a failure
and retry with smaller timesteps (subcycling to make up the original
required timestep).  Among the conditions that can trigger this are
density falling below zero during advection, the ODE integration
failing to converge in the implicit solve, or violation of one of the
timestep limiters during the step.  This means that equations
\ref{eq:dt:nuce} and \ref{eq:dt:nucX} are not reactive, but instead
guaranteed to be met throughout the simulation, because the step is
rejected if they are violated.  The retry mechanism in \castro\ works
with both the Strang and simplified-SDC integration scheme.

@@

We start by writing our update as:
\begin{equation}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left [ \Advs{\Uc} + \Rb(\Uc) \right ] dt
\end{equation}
For the simplified SDC method we explore here, we will make the
advective term piecewise constant in time, using the value at the
midpoint in time, to achieve second-order accuracy, giving us:
\begin{equation}
\label{eq:integral:simplesdc}
\Uc^{n+1} = \Uc^n + \int_t^{t+\Delta t} \left \{ \Adv{\Uc}^{n+1/2} + \Rb(\Uc) \right \} dt
\end{equation}
This is the approach first shown in \citep{SDC-old}.  A
version of this was also implemented in the \maestro\ low-Mach number
hydrodynamics code some time ago.  The compressible version for use in
\castro\ is slightly different due to the need to do some operations
on the conserved variable state and some on the primitive variable
state.  We describe the application of this to the equations of
compressible hydrodynamics below.

@@

  \begin{itemize}
  \item We need an approximation of how much the reactions alone
    changed the primitive variable state over the timestep, which we
    will denote $\Ic_q$.  Since we do not have any information about
    the current timestep in the first iteration,
    we use the value from the last iteration of the previous timestep:
    \begin{equation}
      \Ic^{n+1/2,(-1)}_{\qb} = \Ic^{n-1/2,(\smax-1)}_{\qb}
    \end{equation}
  %% \item Solve the Poisson problem for the initial gravitational potential:
  %%   \begin{equation}
  %%     \nabla^2 \Phi^n = 4\pi G \rho^n
  %%   \end{equation}

@@

    We now seek the $\Ic$'s that capture the effect of just the
    reaction sources on the state variables for the next iteration.
    For the conserved quantities, these would simply be: \MarginPar{fix signs on A}
    \begin{equation}
      \Ic^{(k)}_{\Uc} = \frac{\Uc^{n+1,(k)} - \Uc^n}{\Delta t} - \Adv{\Uc}^{n+1/2,(k)}
    \end{equation}
    However, for our primitive variables, which are used in the
    prediction, we need to construct the required source terms more
    carefully.  We want:
    \begin{equation}
      \label{eq:Iq}
      \Ic^{(k)}_{\qb} = \frac{\qb^{n+1,(k)} - \qb^n}{\Delta t} - \Adv{\qb}^{n+1/2,(k)}
    \end{equation}
    but we need the advective update for $\qb$, which we have not
    constructed.  Additionally, we cannot simply use the equation of
    state on $\Ic^{(k)}_{\Uc}$ since this is a time-derivative and
    does not represent a well-defined state in itself.  Instead, we
    derive $\Ic^{(k)}_{\qb}$ via a multi-step process.  We first find
    the conservative state as if it were updated only with advection:
    \begin{equation}
      \Uc^\star = \Uc^n + \Delta t \Adv{\Uc}^{n+1/2,(k)}
    \end{equation}
    and then construct the corresponding primitive variable state via an algebraic transform,
    $\Uc^\star \rightarrow \qb^\star$.
    This allows us to define the advective update for $\qb$ as:
    \begin{equation}
      \Adv{\qb}^{n+1/2,(k)} = \frac{\qb^\star - \qb^n}{\Delta t}
    \end{equation}
    Defining the primitive state corresponding to the fully-updated
    conserved state via an algebraic transform, $\Uc^{n+1,(k)}
    \rightarrow \qb^{n+1,(k)}$, we can construct $\Ic^{(k)}_{\qb}$ as given
    in Eq.~\ref{eq:Iq}.
    Putting all of this together, we see:
    \begin{equation}
      \Ic^{(k)}_{\qb} = \frac{\qb^{n+1,(k)} - \qb^\star}{\Delta t}
    \end{equation}

@@

For this study we will use the 19 nuclei network containing
\isot{H}{1}, \isot{He}{3}, \isot{He}{4}, \isot{C}{12}, \isot{N}{14},
\isot{O}{16}, \isot{Ne}{20}, \isot{Mg}{24}, \isot{Si}{28},
\isot{S}{32}, \isot{Ar}{36}, \isot{Ca}{40}, \isot{Ti}{44},
\isot{Cr}{48}, \isot{Fe}{52}, \isot{Fe}{54}, \isot{Ni}{56}, protons
(from photodisintegration), and neutrons.  This is based on the
``aprox19'' network from \cite{aprox19} and originally described in
\cite{Kepler}.  We combine this with a table that gives the nuclear
statistical equilibrium (NSE) abundances in regions where the system
is in NSE.  The NSE table was generated using a 127 nuclei reaction
network and is the same as described in \cite{ma:2013}.  In our
simulations, we carry all 19 isotopes in the main network in each zone
and advect them in the hydrodynamics portion of the algorithm.  The
composition of the larger 127 nuclei network is mapped into the 19
isotopes we carry according to table~\ref{table:nuclei}.

The NSE table provides:
\begin{align}
Y_e &= \sum_k \frac{Z_k X_k}{A_k} \\
\bar{A} &= \left [ \sum_k \frac{X_k}{A_k} \right ]^{-1} \\
\frac{B}{A} &= \sum_k \frac{B_k X_k}{A_k}
\end{align}
where $B_k$ is the binding energy of nucleus $k$.  In our simulations,
we store these 3 quantities as auxiliary data that is carried along
with the rest of the fluid state in each zone.

For Strang split coupling of hydro and reactions, $DX_k/Dt = 0$,
and our evolution equations are:
\begin{align}
\frac{DY_e}{Dt} &= \sum_k \frac{Z_k}{A_k} \frac{DX_k}{Dt} = 0 \\
\frac{D}{Dt} \frac{1}{\bar{A}} &= - \frac{1}{\bar{A}^2} \frac{D\bar{A}}{Dt} = \sum_k \frac{1}{A_k} \frac{DX_k}{Dt} = 0 \rightarrow \frac{D\bar{A}}{Dt} = 0 \\
\frac{D}{Dt} \left (\frac{B}{A} \right ) &= \sum_k \frac{B_k}{A_k} \frac{DX_k}{Dt} = 0
\end{align}
Therefore each of these auxiliary equations obeys an advection
equation in the hydro part of the advancement.  In the SDC algorithm,
there will be a reactive source (an $\Ic_q$) for each of these that is
computed in the same manner as above.

 The compositional quantities
it carries, $\bar{A}$ and $Y_e$, are not representable from the 19
isotopes we carry in the main network. For this reason, when we are
using the NSE network, we always take the composition quantities in
the EOS directly from the auxiliary state in each zone instead of
using the $X_k$ directly. Our equation of state needs $\bar{Z}$ in addition
to the auxiliary quantities, which is is easily computed as
\begin{equation}
\bar{Z} = \bar{A} Y_e
\end{equation}

For Strang splitting, the basic flow of using the aprox19 + NSE network
is as follows:
\begin{itemize}
\item {\em Initialize the problem}

  In addition to initializing the mass fractions, $X_k$, we also fill
  the auxiliary composition data, $Y_e$, $\bar{A}$, and $(B/A)$, using
  these $X_k$.
